### 2014-11-17 text/Chapter06.tex
Unified calculus, move future work (Chapter 6)

\subsection{Semi-lattice formulation}
\subsection{Type inference algorithm}

@@

@@

\newcommand{\SH}{ {\textcolor{sclr}{\mathcal{S}}} }
\newcommand{\sempty}{ {\textcolor{sclr}{ \hat{0}}} }
\newcommand{\sunit}{ {\textcolor{sclr}{\hat{1}}} }
\newcommand{\stimes}{ {\textcolor{sclr}{\diamond}} }
\newcommand{\sflat}{\star}

\newcommand{\amerge}{ \textcolor{aclr}{ \mathbin{\rotatebox[origin=c]{-90}{$\ltimes$}} } }
\newcommand{\asplit}{ \textcolor{aclr}{ \mathbin{\rotatebox[origin=c]{90}{$\ltimes$}} } }
\newcommand{\azero}{ \textcolor{aclr}{\bot} }
\newcommand{\aunit}{ \alift{\cunit} } %\textcolor{aclr}{\textnormal{U}} }
\newcommand{\aweak}{ \alift{\czero} } % \textcolor{aclr}{\textnormal{I}} }

\newcommand{\fcons}{ \times } 
\newcommand{\fempty}{\epsilon}

\section{The unified coeffect calculus}

\subsection{Structure of coeffects}

We describe the algebraic structure of coeffects in three
steps. First, we define a \emph{\cclrd{coeffect scalar}} structure 
which defines the basic building blocks of coeffect information;
then we define \emph{\sclrd{coeffect shapes}} which determines how
coeffect scalar values are related to the free-variable context. 
Finally, we define the \emph{\aclrd{coeffect
    algebra}} which consists of shape-indexed coeffect scalar values.

For example, in bounded reuse the coeffect scalar structure
comprised natural numbers $\cclrd{\mathbb{N}}$ with $\cclrd{+}$ and
$\cclrd{\ast}$ operators. The shape for bounded reuse is the length of the
free-variable context and so the coeffect annotation is a vector of matching length.
Finally, the coeffect algebra specifies how vectors are
concatenated and split in abstraction and application.

In the coeffect system of the calculus, contexts are annotated with shape-indexed coeffects (\eg{}, vectors)  as in
$\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau$. However, functions take just a single 
input parameter and so are annotated with scalar coeffect values as in $\sigma \xrightarrow{\cclrd{r}} \tau$.

\paragraph{Coeffect scalar.}
Coeffect scalar structures are equipped with two operations. In bounded reuse, those 
were $\cclrd{\ast}$ for sequencing (in function application) and $\cclrd{+}$ for context 
sharing (in contraction). Additional structure is needed for variable access and sub-coeffecting.

\begin{definition}
\emph{\cclrd{Coeffect scalar}} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ is a set $\C$
together with elements $\cunit,\czero \in \C$, relation $\cleq$ and binary operations 
$\cseq, \cpar$ such that $(\C, \cseq, \cunit)$ and $(\C, \cpar, \czero)$ are monoids
and $(\C, \cleq)$ is a pre-order. The following distributivity axioms are required:
\begin{align*}
(\cclrd{r}\, \cpar\, \cclrd{s}) \;\cseq\; \cclrd{t} & = (\cclrd{r} \,\cseq\, \cclrd{t}) \;\cpar\; (\cclrd{s}\, \cseq\, \cclrd{t}) \\
\cclrd{t} \;\cseq\; (\cclrd{r}\, \cpar\, \cclrd{s}) & = (\cclrd{t} \,\cseq\, \cclrd{r}) \;\cpar\; (\cclrd{t}\, \cseq\, \cclrd{s})
\end{align*}
\end{definition}

\noindent
The operation $\cseq$ must form a monoid with $\cunit$ to guarantee an underlying 
category in the semantics. 
It models sequential composition with variable access
($\cunit$) as the identity. The other element ($\czero$) is used for
variables that are not accessed. The operation $\cpar$ combines coeffects
for contexts used in multiple places (contraction). 
The notation is inspired by the bounded reuse example, which uses coeffect scalar structure
$(\cclrd{\mathbb{N}}, \ast, +, 1,0, \cclrd{\leq})$, but be aware that $\cseq$ and $\cpar$ do
not always mean $\ast$ and $+$.

The context annotations can be viewed as \emph{containers} of scalar coeffects.
For structural coeffects, the container is a vector, while for flat coeffects, it is
a trivial singleton container. We take inspiration from the work of Abbott \emph{et al.}~\cite{types-containers}
which describes containers in terms of \emph{shapes} and a set of \emph{positions} in each shape.

\paragraph{Coeffect shapes.}
The coeffect system is parameterised by a set of shapes $\SH$. A coeffect annotation 
is indexed by a shape $\sclrd{s} \in \SH$ calculated from the shape of the free-variable vector. The correspondence is not necessarily 
bijective. For example, flat coeffect systems have just a single shape $\SH = \{ \ast \}$.

In the coeffect judgment $\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau$, the coeffect annotation
$\aclrd{R}$ is drawn from the set of coeffect scalars $\C$ indexed by the shape of $\Gamma$.
We write $\aclrd{s} = \slift{\Gamma}$ for the shape corresponding to $\Gamma$. We define shapes by a 
\emph{set} of positions and so we can write $\aclrd{R} \in \sclrd{s} \rightarrow \C$ as a
mapping from positions (defined by the shape) to scalar coeffects. We usually write this as 
the exponent $\aclrd{R} \in \C^{\sclrd{s}}$.

The set of shapes is equipped with an operation that combines shapes (when we combine 
variable contexts), an operation that computes shape from the free-variable contexts, 
and two special shapes in $\SH$ representing empty context and singleton context.

\begin{definition}
  A \emph{\sclrd{coeffect shape}} structure $(\SH, \slift{\textnormal{--}}, \stimes,
  \sempty, \sunit)$ comprises a set $\SH$ with a binary operation
  $\stimes$ on $\SH$ for shape composition, a mapping from contexts to
  shapes $\slift{\Gamma} \in \SH$, and elements $\sempty,\sunit \in
  \SH$ such that $(\SH, \stimes, \sempty)$ is a monoid. The 
  elements $\sempty$ and $\sunit$ represent the shapes of empty and
  singleton free-variable contexts respectively.
\end{definition}

\noindent
As said earlier, we use two kinds of shape structures that describe the shape
of vectors and the shape of trivial singleton container:

\begin{itemize}
\item
Structural coeffect shape is defined as $\sclrd{(\mathbb{N}, {|}\textnormal{--}{|}, +, 0, 1)}$.
We treat numbers as sets $0 = \{\}, 1=\{ \emptyset \}, 2=\{ \emptyset, 1\}, 3=\{\emptyset, 1, 2 \} \ldots$ 
(so that a number is a set of positions). The shape mapping ${|}\Gamma{|}$ returns the number
of variables in $\Gamma$. Empty and singleton contexts are annotated with $0$ and $1$, respectively,
and shapes of combined contexts are added so that ${|}\Gamma_1, \Gamma_2{|} = {|}\Gamma_1{|} + {|}\Gamma_2{|}$.
Therefore, a coeffect annotation is a \emph{vector} $\aclrd{R} \in
\C^{\sclrd{n}}$ and assigns a coeffect scalar $\aclrd{R}(i) \in \C$ for each
variable $x_i$ in the context.

\item
Flat coeffect shape is defined as $\sclrd{(\{\sflat \}, \mathsf{const}\,\,\sflat, \diamond, \sflat, \sflat)}$ where
$\sflat\,\diamond\,\sflat = \sflat$ and $\star=\{\emptyset\}$. That is, there is a single shape
$\sflat$ with a single position and all free-variable contexts have the same singleton shape.
Therefore, a coeffect annotation is drawn from $\C^{\sflat}$ which is iso\-morphic to $\C$
and so a coeffect scalar $\cclrd{r}\in \C$ is associated with every free-variable context.
\end{itemize}

\noindent
Using a shape with \emph{no} positions reduces our system to the simply-typed $\lambda$-calculus with no context annotations.
Trees could be used to build a system akin to bunched typing \cite{substruct-bunched}.

\paragraph{Coeffect algebra.}
The coeffect calculus annotates judgments with 
shape-indexed (or, \emph{shaped}) coeffects.
The \emph{\aclrd{coeffect
    algebra}} structure combines a coeffect scalar and coeffect shape structures
to define shaped coeffects and operations for combining these.
In Section~\ref{sec:intro-coeffects}, shaped coeffects were combined
by the tensor $\times$ in structural examples and
$\cup$ in the implicit parameters example. To capture the examples so far and
those described previously~\cite{petricek2013coeffects}, we distinguish two
operators for combining shaped coeffects.

\begin{definition}
  Given a \cclrd{coeffect scalar} $(\C, \cseq, \cpar, \cunit, \czero,
  \cleq)$ and a \sclrd{coeffect shape} $(\SH, \slift{\textnormal{--}}, \stimes,
  \sempty, \sunit)$ a \emph{\aclrd{coeffect algebra}} extends the two
  structures with $(\amerge, \asplit, \azero)$ where $\azero \in
  \C^{\sempty}$ is a coeffect annotation for the empty context and
  $\amerge, \asplit$ are families of operations that combine coeffect
  annotations indexed by shapes. That is $\forall \sclrd{n},\sclrd{m}
  \in \SH$:
\[
\asplit_{\sclrd{m},\sclrd{n}}, \amerge_{\sclrd{m},\sclrd{n}} ~:~ \C^{\sclrd{m}} \times \C^{\sclrd{n}} \rightarrow \C^{\sclrd{m}\,\stimes\,\sclrd{n}}
\]
\end{definition}

\noindent
A coeffect algebra induces the following two additional operations:
\[
\begin{array}{rclrcl}
\alift{\textnormal{--}} &\narrow{:}& \C \rightarrow \C^{\sunit}          \quad\quad
   & \aseq_{\sclrd{m}} &\narrow{:}&  \C \times \C^{\sclrd{m}} \rightarrow \C^{\sclrd{m}}  
\\ 
\alift{\cclrd{x}} &\narrow{=}& \lambda \sunit . \cclrd{x}  
   & \cclrd{r}\ \aseq\ \aclrd{S} &\narrow{=}& \lambda \sclrd{s} . \cclrd{r}\ \cseq\ (\aclrd{S}(\sclrd{s}))
\end{array}
\]
$\alift{\textnormal{--}}$ lifts a scalar coeffect 
to a shaped coeffect indexed by the singleton context shape. 
The $\aseq_{\sclrd{m}}$ operation is a left multiplication of a vector by scalar.
As we always use lower-case for scalars and upper-case for vectors, using the same
symbol is not ambiguous. We also tend to omit the subscript
$\sclrd{m}$ and write just $\aseq$.

The operators $\asplit$ and $\amerge$ combine shaped
coeffects associated with two contexts. For example,
assume we have $\Gamma_1$ and $\Gamma_2$ with coeffects 
$\aclrd{R} \in \C^{\sclrd{m}}$ and $\aclrd{S} \in \C^{\sclrd{n}}$. In
the structural system, the context shapes $\sclrd{m}, \sclrd{n}$
denote the number of variables in the two contexts. The combined
context $\Gamma_1, \Gamma_2$ has a shape $\sclrd{m}\,\stimes\,\sclrd{n}$ and
the combined coeffects $\aclrd{R} \,\amerge\, \aclrd{S}, \aclrd{R} \,\asplit\, \aclrd{S} \in \C^{\sclrd{m} \stimes \sclrd{n}}$
are indexed by that shape.

For structural coeffect systems such as bounded reuse, both $\asplit$ and $\amerge$ are
just the tensor product $\times$ of vectors. However, we need to distinguish them for flat coeffect 
systems discussed later.

The difference is explained by the semantics (Section~\ref{sec:semantics}), where
$\aclrd{R}\,\amerge\,\aclrd{S}$ is an annotation of the codomain of 
a morphism that merges the capabilities provided by two contexts (in the syntactic 
reading, splits the context requirements); $\aclrd{R}\,\asplit\,\aclrd{S}$ is an annotation of
the domain of a morphism that splits the capabilities of a single context into two parts (in the
syntactic reading, merges their context requi\-rements). Syntactically, this means that we always use
$\amerge$ in the rule \emph{assumptions} and $\asplit$ in \emph{conclusions}. For now, it suffices to use the 
bounded reuse intuition and read the operations as tensor products.

@@

The typing of let-binding (\emph{let}) corresponds to the typing of an expression $(\lambda x.e_2)~e_1$.
Syntactically, the context requirements are first split using $\amerge$ and then re-combined
using $\asplit$.

@@

Exchange and contraction decompose and reconstruct coeffect annotations 
using $\amerge_{\sclrd{m},\sclrd{n}}$ (in assumption) and $\asplit_{\sclrd{m},\sclrd{n}}$ (in conclusion). 
The shape subscripts are omitted, but we require the shapes to match using $\sclrd{m} = \slift{\Gamma_1}$
and $\sclrd{n} = \slift{\Gamma_2}$.

@@

In addition to the coeffect scalar structure, we also need to define $\asplit$ and $\amerge$.
Our examples of flat coeffects use $\cpar$ (merging of scalar coeffects) for $\asplit$ (merging of shaped 
coeffect annotations). However, the $\amerge$ operation needs to be provided explicitly.

>>>>

The main goal of this thesis is to provide a \emph{unified} calculus for tracking context 
dependence. We have not achieved this goal yet. In Chapter~\ref{ch:applications}, we identified 
two kinds of contextual properties that we further covered separately -- flat coeffects in 
Chapter~\ref{ch:flat} track whole-context properties and structural coeffects, covered in 
Chapter~\ref{ch:structural}, track per-variable properties. In this chapter, we unify
the two notions. We introduce a \emph{unified coeffect} calculus that generalizes the two systems
and can be instantiated to track both flat and structural properties (Section~\ref{sec:unified-unified}).

Although the results presented in this thesis are mainly of a theoretical nature, we indeed 
believe that coeffects should be integrated in main-stream programming languages. In the second
part of this chapter, we outline one possible approach for practical implementations of coeffects 
(Section~\ref{sec:unified-impl}). Finally, we discuss an alternative approach to defining coeffect 
systems which highlights the relationship between our work and related work arising from modal logics 
(Section~\ref{sec:unified-meta}).

@@

\section{The unified coeffect calculus}
\label{sec:unified-unified}

The flat coeffect calculus (Figure~\ref{fig:flat-types}) and the structural coeffect calculus 
differ in a number of ways (Figure~\ref{fig:struct-types}). Understanding the differences is the 
key to reconciling the two systems:

\begin{itemize}
\item Structural coeffect calculus contains explicit rules for context manipulation
  (weakening, contraction, exchange). In flat coeffect calculus, these rules are not defined
  explicitly, but are admissible.

\item In structural coeffect calculus, the variable context is treated as a vector
  and is annotated with a vector of (scalar) coeffects. In flat coeffect calculus,
  the variable context is a set and is annotated with a single (scalar) coeffect.

\item In flat coeffect calculus, we distinguish between splitting of the context requirements
  and merging of context requirements ($\cpar$ and $\czip$, respectively). In structural
  coeffect calculus, the operations (which model splitting and appending vectors) are invertible 
  and so the structural coeffect algebra requires just $\atimes$.
\end{itemize}

\noindent
In the unified calculus presented in this section, we address the three differences as follows.
We use calculus with explicit rules for context manipulation. In systems that arise from flat
calculi, the rules can be applied freely without changing the coeffects. We generalize the 
structure of coeffect annotations using the notion of a ``container'' which can be specialized 
to obtain a single annotation or a vector of annotations. Finally, we distinguish between 
splitting and merging of context requirements (using the notation $\apar$ and $\azip$, 
respectively). For structural coeffect calculi, the two operators coincide, but for flat
coeffect calculi, they provide the needed flexibility.

\subsection{Shapes and containers}

Our notion of a \emph{coeffect container} is based on the idea of a container introduced by 
Abbott et al. \cite{types-containers}. Interestingly, the work on containers has later been linked
to comonads by Ahman et al. \cite{comonads-containers}. Intuitively a container describes data
types such as lists, trees or streams. A container is formed by shapes (\eg~lengths of lists).
For every shape, we can obtain a set of positions in the container (\eg~offsets in a list of a
specified length). More formally:

\begin{definition}
A \emph{container} $S \triangleleft P$ is given by a set $S$ of shapes and a shape-indexed family
$P : S \rightarrow \ident{Set}$ of positions.
\end{definition}

\noindent
Well-known examples of containers include lists, non-empty lists, (un\-boun\-ded) streams and 
singleton data type (which contains exactly one element). Two containers relevant to our work
are lists and singleton data types:

\begin{itemize}
\item The container representing lists is given by $S \triangleleft P$ where 
  shapes are integers $S = \ident{Nat}$ (lengths of a list). The set of positions for a 
  given length $n$ is a set of indices $P n = \{ 1 \ldots n \}$.

\item The contianer representing singleton data type is given by $S \triangleleft P$ where
  shapes are given by a singleton set $S = \{ \ast \}$ and the set of positions for the
  shape $\ast$ contains exactly one position $P \ast = \{ 0 \}$.
\end{itemize}

\noindent
In the unified coeffect calculus, the structure of coeffect annotations is defined by a 
container with additional operations (discussed later) that links it with the free-variable 
context $\Gamma$.

\subsection{Structure of coeffects}
In structural coeffect calculus, the structure annotation was formed by a vector of coeffect
sclars. The unified coeffect calculus is similar, but a \emph{vector} is replaced with a 
\emph{container}. The primitive coeffect annotations in the unified calculus are formed by
a \emph{coeffect scalar}, which remains the same as in structural coeffect calculus
(Definition~\ref{def:structural-scalar}). In this section, we refer to it as
\emph{\cclrd{unified coeffect scalar}} (and we repeat the definition below). Then we
define \emph{\sclrd{unified coeffect containers}} which determines how coeffect scalar values 
are attached to the free-variable context. Finally, we define the \emph{\aclrd{unified coeffect 
algebra}} which consists of shape-indexed coeffect scalar values.

As in the structural coeffect calculus, the contexts in the unified calculus are annotated with 
shape-indexed coeffects, written as $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$; 
functions take just a single input parameter and so are annotated with scalar coeffect values 
$\sigma \xrightarrow{\cclrd{r}} \tau$.

\paragraph{Coeffect scalar.}
The following definition of the coeffect scalar structure repeats the Definition~\ref{def:structural-scalar}
from the previous chapter.

\begin{definition}
A \emph{\cclrd{unified coeffect scalar}} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ is a set 
$\C$ together with elements $\cunit, \czero \in \C$, relation $\cleq$ and binary operations 
$\cseq, \cpar$ such that $(\C, \cseq, \cunit)$ and $(\C, \cpar, \czero)$ are monoids and
$(\C, \cleq)$ is a pre-order. That is, for all $r,s,t\in \C$:
%
\begin{equation*}
\begin{array}{ccr}
r \;\cseq\; (s \;\cseq\; t) = (r \;\cseq\; s) \;\cseq\; t  &
\cunit \;\cseq\; r = r = r \;\cseq\; \cunit &
\textnormal{(monoid)}   
\\
r \;\cpar\; (s \;\cpar\; t) = (r \;\cpar\; s) \;\cpar\; t &
\czero \;\cpar\; r = r = r \;\cpar\; \czero &
\textnormal{(monoid)}   
\\
\textnormal{if}~~r\; \cleq\; s ~~\textnormal{and}~~s\; \cleq\; t~~\textnormal{then}~~r\; \cleq\; t&
t\; \cleq\; t &
\textnormal{(pre-order)}   
\end{array}
\end{equation*}
\end{definition}

\noindent
As previously, the monoid $(\C,\cseq,\cunit)$ models sequential composition; the laws guarantee
an underlying category structre; $\cunit$ and $\czero$ represent an accessed and unused variable, 
respectively.

The $\cpar$ operation models combining of context requirements arising from multiple parts of a 
program. The meaning depends on the coeffect container. The operation can either combine requirements
of individual variables (structural) or requirements attached to the whole context of multiple
sub-exp\-ressions (flat).

\paragraph{Coeffect containers.}
The coeffect container is a container that determines how are scalar coeffect annotations attached 
to free-variable contexts. In addition to a container $\SHP$ with shapes and shape-indexed positions, the
coeffect container also provides a mapping that returns the shape of a free-variable context. The
mapping between the shape of the variable context and the shape of the coeffect annotation
is not necessarily bijective. For example, coeffect annotations in flat systems have just a single 
shape $\SH = \{ \ast \}$.

In the coeffect judgment $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$, the coeffect annotation
$\aclrd{\textbf{r}}$ is drawn from the set of coeffect scalars $\C$ indexed by the shape of $\Gamma$.
We write $\sclrd{s} = \slen{\Gamma}$ for the shape corresponding to $\Gamma$. The operation $\SP\sclrd{s}$
returns a \emph{set} of positions and so we can write $\aclrd{\textbf{r}} \in \SP\sclrd{s} \rightarrow \C$ 
as a mapping from positions (defined by the shape) to scalar coeffects. We usually write this as 
the exponent $\aclrd{\textbf{r}} \in \C^{\SP\sclrd{s}}$.

The coeffect container is also equipped with an operation that appends shapes (when we concatenate
variable contexts) and two special shapes in $\SH$ representing empty context and singleton context.

\begin{definition}
A \emph{\sclrd{coeffect container}} structure $(\SHP, \stimes, \sempty, \sunit, \slen{-})$ 
comprises a container $\SHP$ with a binary operation $\stimes$ on $\SH$ for appending shapes, a 
mapping from free-variable contexts to shapes $\slen{\Gamma} \in \SH$, and elements $\sempty,\sunit \in
\SH$ such that $(\SH, \stimes, \sempty)$ is a monoid.

The elements $\sempty$ and $\sunit$ represent the shapes of empty and singleton free-variable contexts 
respectively. The $\stimes$ operation corresponds to concatentation of free-variable contexts. Given
$\Gamma_1$ and $\Gamma_2$ such that $\sclrd{s_1}=\slen{\Gamma_1}, \sclrd{s_2}=\slen{\Gamma_2}$, 
we require that $\sclrd{s_1}\,\stimes\,\sclrd{s_2}=\slen{\Gamma_1, \Gamma_2}$. 
\end{definition}

\noindent
As said earlier, we use two kinds of coeffect containers that describe the structure of vectors 
(for structural coeffects) and the shape of trivial singleton container (for flat coeffects):

\begin{example}
Structural coeffect shape is defined as $\sclrd{(\SHP, {|}\textnormal{--}{|}, +, 0, 1)}$
where $\SH=\mathbb{N}$ and $\SP \sclrd{n}=\{ 1 \ldots n \}$. The shape mapping ${|}\Gamma{|}$ returns the 
number of variables in $\Gamma$. Empty and singleton contexts are annotated with $0$ and $1$, 
respectively, and shapes of combined contexts are added so that ${|}\Gamma_1, \Gamma_2{|} = 
{|}\Gamma_1{|} + {|}\Gamma_2{|}$.

Therefore, a coeffect annotation is a \emph{vector} 
$\aclrd{\textbf{r}} \in \C^{\SP\sclrd{n}}$ and assigns a coeffect scalar $\aclrd{\textbf{r}}(i) \in \C$ 
for each position (corresponding to a variable $x_i$ in the context).
\end{example}

\begin{example}
Flat coeffect shape is defined as $\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$.
The container is defined as a singleton data type $\SH = \{\sflat \}$ and $\SP \sflat = \{ 0 \}$
with a constant function ${|}\Gamma{|}=\sflat$ and a trivial operation $\sflat\,\diamond\,\sflat = \sflat$.

That is, there is a single shape
$\sflat$ with a single position and all free-variable contexts have the same singleton shape.
Therefore, a coeffect annotation is drawn from $\C^\{\sflat\}$ which is isomorphic to $\C$
and so a coeffect scalar $\cclrd{r}\in \C$ is associated with every free-variable context.
\end{example}

\newpage
\begin{example}
Similarly to the previous example, we can also define a coeffect container with \emph{no} positions,
\ie~$\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$ where $\SH = \{\sflat \}$, $\SP \sflat = \emptyset$,
${|}\Gamma{|}=\sflat$ and $\sflat\,\diamond\,\sflat = \sflat$.

This reduces our system to the simply-typed $\lambda$-calculus with no context annotations, because
$\SP\sflat=\emptyset$ and so coeffect annotations would be from the set $\C^\emptyset$.
\end{example}

\paragraph{Unified coeffect algebra.}
The coeffect calculus annotates judgments with shape-indexed coeffect annotations.
The \emph{\aclrd{unified coeffect algebra}} combines a coeffect scalar and coeffect container
to define shape-indexed coeffects and operations for manipulating these.

The definition here reconciles the third point discussed in Section~\ref{sec:unified-unified} --
the fact that flat coeffects use separate operations for splitting and merging ($\cpar$ and $\czip$)
while structural coeffects use tensor $\atimes$. In the unified calculus, we use two operators
that can, however, be coincide.

\begin{definition}
Given a \cclrd{unified coeffect scalar} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and a 
\sclrd{coeffect container} $(\SHP, \slen{-}, \stimes, \sempty, \sunit)$ a \emph{\aclrd{unified 
coeffect algebra}} extends the two structures with $(\azip, \apar, \azero)$ where $\azero$ is a 
coeffect annotation for the empty context and $\azip, \apar$ are families of operations that 
combine coeffect annotations indexed by shapes. That is $\forall \sclrd{n},\sclrd{m} \in \SH$:
%
\begin{equation*}
\apar_{\sclrd{m},\sclrd{n}}, \azip_{\sclrd{m},\sclrd{n}} ~:~ \C^{\SP \sclrd{m}} \times \C^{\SP \sclrd{n}} \rightarrow \C^{\SP (\sclrd{m}\,\stimes\,\sclrd{n})}
\end{equation*}
\end{definition}

\noindent
A coeffect algebra induces the following three additional operations:
%
\begin{equation*}
\begin{array}{rclcrcl}
 \alift{-} &\narrow{:}& \C \rightarrow \C^{\SP \sunit}  \\
 \alift{\cclrd{x}} &\narrow{=}& \lambda \sunit . \cclrd{x}   \\
\\[-0.5em]
 \aseq_{\sclrd{m}} &\narrow{:}&  \C \times \C^{\SP \sclrd{m}} \rightarrow \C^{\SP \sclrd{m}}   \\
 \cclrd{r}\, \aseq\, \aclrd{\textbf{s}} &\narrow{=}& \lambda \sclrd{s} . \cclrd{r}\, \cseq\, \aclrd{\textbf{s}}(\sclrd{s}) \\
\\[-0.5em]
 \alen{-} &\narrow{:}& \C^{\SP \sclrd{m}} \rightarrow \sclrd{m}\\
 \alen{\aclrd{\textbf{r}}} &\narrow{=}& \sclrd{m}
\end{array}
\end{equation*}
%
The $\alift{-}$ operation lifts a scalar coeffect to a shape-indexed coeffect that is indexed by 
the singleton context shape. The $\aseq_{\sclrd{m}}$ operation is a left multiplication of a 
vector by scalar. As we always use bold font for vectors and ordinary font for scalars (as well 
as distinct colour), using the same symbol is not ambiguous. We also tend to omit the subscript
$\sclrd{m}$ and write $\aseq$.

Finally, we define $\alen{}$ as an operation that returns the shape of a given shape-indexed 
coeffect. The only purpose is to simplify notation, as we often need to specify that shapes of 
variable context and coeffect match, \eg~$\alen{\aclrd{\textbf{r}}} = \slen{\Gamma}$.

\paragraph{Splitting and merging coeffects.}
The operators $\apar$ and $\azip$ combine shape-indexed coeffects associated with two contexts. 
For example, assume we have $\Gamma_1$ and $\Gamma_2$ with coeffects $\aclrd{\textbf{r}} \in 
\C^{\SP\sclrd{m}}$ and $\aclrd{\textbf{s}} \in \C^{\SP\sclrd{n}}$. In the structural system, the 
context shapes $\sclrd{m}, \sclrd{n}$ denote the number of variables in the two contexts. The 
combined context $\Gamma_1, \Gamma_2$ has a shape $\sclrd{m}\,\stimes\,\sclrd{n}$ and the combined 
coeffects $\aclrd{\textbf{r}} \,\azip\, \aclrd{\textbf{s}}, \aclrd{\textbf{r}} \,\apar\, \aclrd{\textbf{s}} 
\in \C^{\SP(\sclrd{m} \stimes \sclrd{n})}$ are indexed by that shape.

For structural coeffect systems such as bounded reuse, both $\apar$ and $\azip$ are
just the tensor product $\times$ of vectors. For flat coeffect systems, the operations
can be defined independently, letting $\azip=\czip$ and $\apar=\cpar$.

The difference between $\azip$ and $\apar$ is clarified by the semantics (Sec\-tion~\ref{sec:unified-semantics}), 
where $\aclrd{\textbf{r}}\,\azip\,\aclrd{\textbf{s}}$ is an annotation of the \emph{codomain} of 
a morphism that merges the capabilities provided by two contexts (in the syntactic 
reading, splits the context requirements), while $\aclrd{\textbf{r}}\,\apar\,\aclrd{\textbf{s}}$ 
is an annotation of the \emph{domain} of a morphism that splits the capabilities of a single 
context into two parts (in the syntactic reading, merges their context requirements). 
Syntactically, this means that we always use $\azip$ in the rule \emph{assumptions} and 
$\apar$ in \emph{conclusions}.

@@

\begin{figure}[t]

{\small a.) Syntax-driven typing rules:}
\begin{equation*}
\tyrule{var}
  { }
  {\coctx{x\!:\!\tau}{\alift{\cunit}} \vdash x : \tau }
\end{equation*}
\begin{equation*}
\tyrule{const}
  {c \!:\! \tau \in \Delta}
  {\coctx{()}{\azero} \vdash c : \tau }
\end{equation*}
\begin{equation*}
\tyrule{app}
  {\coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 &
   \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1 }
  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}\;\apar\;(\cclrd{t} \,\aseq\, \aclrd{\textbf{s}})}} \vdash e_1~e_2 : \tau_2}
\end{equation*}
\begin{equation*}
\tyrule{abs}
  {\coctx{\Gamma, x\!:\!\tau_1}{\aclrd{\textbf{r}}\,\azip\,\alift{\cclrd{s}}} \vdash e : \tau_2}
  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
\end{equation*}
\begin{equation*}
\tyrule{let}
  { \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 &
    \coctx{\Gamma_2, x\!:\!\tau_1}{\aclrd{\textbf{s}}\;\azip\;\alift{\cclrd{t}}} \vdash e_2 : \tau_2}
  {\coctx{\Gamma_1, \Gamma_2}{(\cclrd{t} \,\aseq\, \aclrd{\textbf{r}}) \;\apar\; \aclrd{\textbf{s}}} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
\end{equation*}

\vspace{1em}
{\small b.) Structural rules for context manipulation:}
\begin{equation*}
\tyrule{sub}
  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \azip \alift{\cclrd{s'}} \azip \aclrd{\textbf{q}}} \vdash e : \tau}
  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \apar \alift{\cclrd{s}} \apar \aclrd{\textbf{q}}} \vdash e : \tau}
\quad
(\cclrd{s'} \cleq \cclrd{s})
\end{equation*}
\begin{equation*}
\tyrule{weak}
  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \tau}
  {\coctx{\Gamma,x \!:\! \tau_1}{\aclrd{\textbf{r}} \apar \alift{ \czero }} \vdash e : \tau} 
\end{equation*}
\begin{equation*}
\tyrule{exch}
  {\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\aclrd{\textbf{r}} \azip \alift{\cclrd{s}} \azip \alift{\cclrd{t}} \azip \aclrd{\textbf{q}}} \vdash e : \tau}
  {\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \apar \alift{\cclrd{t}} \apar \alift{\cclrd{s}} \apar \aclrd{\textbf{q}}} \vdash e : \tau}
\quad
\begin{array}{l}
 \slen{\Gamma_1} = \alen{\aclrd{\textbf{r}}}\\[-0.25em]
 \slen{\Gamma_2} = \alen{\aclrd{\textbf{s}}}
\end{array}
\end{equation*}
\begin{equation*}
\hspace{2.5em} 
\tyrule{contr}
  {\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \azip \alift{\cclrd{s}} \azip \alift{\cclrd{t}} \azip \aclrd{\textbf{q}}} \vdash e : \tau}
  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \apar \alift{\cclrd{s} \,\cpar\, \cclrd{t}} \apar \aclrd{\textbf{q}}} \vdash \subst{e}{z,y}{x} : \tau}
~
\begin{array}{l}
 \slen{\Gamma_1} = \alen{\aclrd{\textbf{r}}}\\[-0.25em]
 \slen{\Gamma_2} = \alen{\aclrd{\textbf{s}}}
\end{array}
\end{equation*}

\caption{Type system for the unified coeffect calculus}
\label{fig:unif-types}
\end{figure}

@@

The typing of let-binding (\emph{let}) corresponds to the typing of an expression $(\lambda x.e_2)~e_1$.
Syntactically, the context requirements are first split using $\azip$ and then re-combined
using $\apar$.

@@

Exchange and contraction decompose and reconstruct coeffect annotations 
using $\azip_{\sclrd{m},\sclrd{n}}$ (in assumption) and $\apar_{\sclrd{m},\sclrd{n}}$ (in conclusion). 
The shape subscripts are omitted, but we require the shapes to match using $\sclrd{m} = \slen{\Gamma_1}$
and $\sclrd{n} = \slen{\Gamma_2}$.

@@

In addition to the coeffect scalar structure, we also need to define $\apar$ and $\azip$.
Our examples of flat coeffects use $\cpar$ (merging of scalar coeffects) for $\apar$ (merging of shaped 
coeffect annotations). However, the $\azip$ operation needs to be provided explicitly.
